\thetitle

### 2025-06-26 workshop/PT.tex
replace PT appendix with nutshell section

\begin{figure}[t]
    \centering
    \includegraphics[width=0.25\textwidth]{../img/bimodal.pdf}
    \caption{A simple bimodal distribution from which traditional 
    MCMC methods may struggle to obtain samples. Blue lines display output 
    from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
    The sampler in this figure is visibly trapped in one of the two modes.}
    \label{fig:bimodal}
\end{figure}

@@

% NB: this is placed here just to make it appear in a different page
\begin{figure}[t]
    \centering
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_1.pdf}
      \caption*{$\pi_1$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_2.pdf}
      \caption*{$\pi_2$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_3.pdf}
      \caption*{$\pi_3$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_4.pdf}
      \caption*{$\pi_4$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_5.pdf}
      \caption*{$\pi_5$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_6.pdf}
      \caption*{$\pi_6$}
    \end{minipage}
    \caption{Heatmaps of six distributions lying on an annealing path 
    from a unimodal reference distribution  
    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
    $\pi_N$, which is in this case the bimodal distribution from 
    \cref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
    are not directly comparable because the densities of intermediate distributions 
    are not normalized.)}
    \label{fig:path}
\end{figure}

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=0.33\textwidth]{../img/speedup_distributed.pdf}
    \hspace{4em}
    \includegraphics[width=0.33\textwidth]{../img/speedup_parallel.pdf}
    \caption{
        Weak scaling results for PT in both the distributed (MPI, {\bf left}) 
        and parallel (Julia threads, {\bf right}) settings 
        with approximate 95\% confidence intervals for the speedup relative to single-thread execution.   
        The blue dotted line represents a hypothetical speedup equal to the 
        number of processes/threads. 
        The red dashed line indicates a speedup of one for reference.
    }
    \label{fig:PT_scaling}
\end{figure*}

>>>>

%\begin{figure}[t]
%    \centering
%    \includegraphics[width=0.25\textwidth]{../img/bimodal.pdf}
%    \caption{A simple bimodal distribution from which traditional 
%    MCMC methods may struggle to obtain samples. Blue lines display output 
%    from 1,000 iterations of a Metropolis-Hastings random walk MCMC algorithm.
%    The sampler in this figure is visibly trapped in one of the two modes.}
%    \label{fig:bimodal}
%\end{figure}

@@

%% NB: this is placed here just to make it appear in a different page
%\begin{figure}[t]
%    \centering
%    \begin{minipage}{0.15\textwidth}
%      \centering
%      \includegraphics[width=\textwidth]{../img/heatmap_path_1.pdf}
%      \caption*{$\pi_1$}
%    \end{minipage}
%    \begin{minipage}{0.15\textwidth}
%      \centering
%      \includegraphics[width=\textwidth]{../img/heatmap_path_2.pdf}
%      \caption*{$\pi_2$}
%    \end{minipage}
%    \begin{minipage}{0.15\textwidth}
%      \centering
%      \includegraphics[width=\textwidth]{../img/heatmap_path_3.pdf}
%      \caption*{$\pi_3$}
%    \end{minipage}
%    \begin{minipage}{0.15\textwidth}
%      \centering
%      \includegraphics[width=\textwidth]{../img/heatmap_path_4.pdf}
%      \caption*{$\pi_4$}
%    \end{minipage}
%    \begin{minipage}{0.15\textwidth}
%      \centering
%      \includegraphics[width=\textwidth]{../img/heatmap_path_5.pdf}
%      \caption*{$\pi_5$}
%    \end{minipage}
%    \begin{minipage}{0.15\textwidth}
%      \centering
%      \includegraphics[width=\textwidth]{../img/heatmap_path_6.pdf}
%      \caption*{$\pi_6$}
%    \end{minipage}
%    \caption{Heatmaps of six distributions lying on an annealing path 
%    from a unimodal reference distribution  
%    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
%    $\pi_N$, which is in this case the bimodal distribution from 
%    \cref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
%    are not directly comparable because the densities of intermediate distributions 
%    are not normalized.)}
%    \label{fig:path}
%\end{figure}

@@

%\begin{figure*}[t]
%    \centering
%    \includegraphics[width=0.33\textwidth]{../img/speedup_distributed.pdf}
%    \hspace{4em}
%    \includegraphics[width=0.33\textwidth]{../img/speedup_parallel.pdf}
%    \caption{
%        Weak scaling results for PT in both the distributed (MPI, {\bf left}) 
%        and parallel (Julia threads, {\bf right}) settings 
%        with approximate 95\% confidence intervals for the speedup relative to single-thread execution.   
%        The blue dotted line represents a hypothetical speedup equal to the 
%        number of processes/threads. 
%        The red dashed line indicates a speedup of one for reference.
%    }
%    \label{fig:PT_scaling}
%\end{figure*}

### 2025-06-26 workshop/SPI.tex
acc + fix tree lengths
replace PT appendix with nutshell section

\section{Strong parallelism invariance}
\label{sec:PI_causes}
In this section we describe potential violations of SPI that can occur in a 
distributed setting. We also explain how we avoid these issues by using special 
distributed reduction schemes and splittable random number generators. 
Insights provided in this section can be applied 
to general distributed software beyond Julia.

@@

\subsection{Distributed reduction and floating point non-associativity}
The execution of PT in Pigeons.jl is divided into \emph{rounds}, which are
sequences of \emph{scans}---a single call to \cref{alg:distributed_PT}---of 
exponentially increasing length. At the end of each round,
\emph{replicas}---the fundamental unit of parallelization in PT (see \cref{sec:PT} for 
details)---need to exchange information. This can be related to the output of the 
program---e.g., the value of the normalization constant $Z$---as well as statistics 
needed to adapt the sampler so that the next round runs more efficiently.
Almost all of these cases involve summing floating-point numbers that are 
located on different threads, processes, or compute nodes (in the following we use 
the generic term \emph{machine}).

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}
        [{$10x$}
          [{$3x$}
            [{$1x$}]
            [{$2x$}]
          ]
          [{$7x$}
            [{$3x$}] 
            [{$4x$}] 
          ]
        ]
        [{$26x$}
          [{$11x$}
            [{$5x$}] 
            [{$6x$}] 
          ]
          [{$15x$}
            [{$7x$}] 
            [{$8x$}] 
          ]
        ]
      ]
  \end{forest}
  \caption{Adding eight floating point numbers $\cbra{1x, 2x, \ldots, 8x}$ 
  across eight machines. 
  Additions in each row of the tree can be performed in parallel. 
  The final result is stored in the root node of the tree and can be represented 
  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$, indicating 
  the order of operations.}
  \label{fig:reduction_tree_8}
\end{figure}

\begin{figure}[t]
  \centering
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}
        [{$10x$}
          [{$6x$}
            [{$3x$}
              [{$1x$}]
              [{$2x$}]
            ]
            [{$3x$}]
          ]
          [{$4x$}]
        ]
        [{$26x$}
          [{$8x$}]
          [{$18x$}
            [{$7x$}]
            [{$11x$}
              [{$6x$}]
              [{$5x$}]
            ]
          ]
        ]
      ]
  \end{forest}
  \caption{One possible way of adding eight floating point numbers 
  $\cbra{1x, 2x, \ldots, 8x}$ across two machines. 
  The final result is stored in the root node of the tree and can be represented 
  by the expression $(((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x)$.
  Note that the order of operations in this expression is different from the one 
  presented in \cref{fig:reduction_tree_8}.}
  \label{fig:reduction_tree_2}
\end{figure}

@@

\begin{figure*}[t]
  \centering
  \begin{minipage}{0.5\linewidth}
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}, fill = julia1
        [{$10x$}, fill = julia1
          [{$3x$}, fill = julia1
            [{$1x$}, fill = julia1]
            [{$2x$}, fill = julia2]
          ]
          [{$7x$}, fill = julia3
            [{$3x$}, fill = julia3] 
            [{$4x$}, fill = julia4] 
          ]
        ]
        [{$26x$}, fill = julia5
          [{$11x$}, fill = julia5
            [{$5x$}, fill = julia5] 
            [{$6x$}, fill = julia6] 
          ]
          [{$15x$}, fill = julia7
            [{$7x$}, fill = julia7] 
            [{$8x$}, fill = julia8] 
          ]
        ]
      ]
  \end{forest}
  \end{minipage}%
  \begin{minipage}{0.5\linewidth}
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}, fill = julia1
        [{$10x$}, fill = julia1
          [{$3x$}, fill = julia1
            [{$1x$}, fill = julia1]
            [{$2x$}, fill = julia1]
          ]
          [{$7x$}, fill = julia1
            [{$3x$}, fill = julia1] 
            [{$4x$}, fill = julia1] 
          ]
        ]
        [{$26x$}, fill = julia2
          [{$11x$}, fill = julia2
            [{$5x$}, fill = julia2] 
            [{$6x$}, fill = julia2] 
          ]
          [{$15x$}, fill = julia2
            [{$7x$}, fill = julia2] 
            [{$8x$}, fill = julia2] 
          ]
        ]
      ]
  \end{forest}
  \end{minipage}
  \caption{Addition of eight floating point numbers across eight (\textbf{left}) 
  and two (\textbf{right}) machines with a guarantee on SPI. Each machine is 
  represented by a different colour. In both cases, the final result can be 
  represented by the same expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$.}
  \label{fig:reduction_tree_colour}
\end{figure*}

>>>>

\section{Strong parallelism invariance}\label{sec:PI_causes}

In this section we describe potential violations of \emph{strong 
parallelism invariance} (SPI)---which we define as a guarantee that the output 
for a given seed is identical irrespective of the number of threads and 
processes---that can occur in a distributed setting. We also explain how we 
avoid these issues by using special distributed reduction schemes and 
splittable random number generators. Insights provided in this section 
can be applied to general distributed software beyond Julia.

@@

\subsection{Distributed reduction and floating point non-associativity}
The execution of PT in Pigeons.jl is divided into \emph{rounds}, which are
sequences of \emph{scans}---an exploration phase followed by a communication phase---%
of exponentially increasing length. At the end of each round,
\emph{replicas}---the fundamental unit of parallelization in PT---need to exchange information. This can be related to the output of the 
program---e.g., the value of the normalization constant $Z$---as well as statistics 
needed to adapt the sampler so that the next round runs more efficiently.
Almost all of these cases involve summing floating-point numbers that are 
located on different threads, processes, or compute nodes (in the following we use 
the generic term \emph{machine}).

\begin{figure}[t]
  \footnotesize
  \centering
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}
        [{$10x$}
          [{$3x$}
            [{$1x$}]
            [{$2x$}]
          ]
          [{$7x$}
            [{$3x$}] 
            [{$4x$}] 
          ]
        ]
        [{$26x$}
          [{$11x$}
            [{$5x$}] 
            [{$6x$}] 
          ]
          [{$15x$}
            [{$7x$}] 
            [{$8x$}] 
          ]
        ]
      ]
  \end{forest}
  \caption{Adding eight floating point numbers $\cbra{1x, 2x, \ldots, 8x}$ 
  across eight machines. 
  Additions in each row of the tree can be performed in parallel. 
  The final result is stored in the root node of the tree and can be represented 
  by the expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$, indicating 
  the order of operations.}
  \label{fig:reduction_tree_8}
\end{figure}

\begin{figure}[t]
  \footnotesize
  \centering
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}
        [{$10x$}
          [{$6x$}
            [{$3x$}
              [{$1x$}]
              [{$2x$}]
            ]
            [{$3x$}]
          ]
          [{$4x$}]
        ]
        [{$26x$}
          [{$8x$}]
          [{$18x$}
            [{$7x$}]
            [{$11x$}
              [{$6x$}]
              [{$5x$}]
            ]
          ]
        ]
      ]
  \end{forest}
  \caption{One possible way of adding eight floating point numbers 
  $\cbra{1x, 2x, \ldots, 8x}$ across two machines. 
  The final result is stored in the root node of the tree and can be represented 
  by the expression $(((1x + 2x) + 3x) + 4x) + (((5x + 6x) + 7x) + 8x)$.
  Note that the order of operations in this expression is different from the one 
  presented in \cref{fig:reduction_tree_8}.}
  \label{fig:reduction_tree_2}
\end{figure}

@@

\begin{figure*}[t]
  \footnotesize
  \centering
  \begin{minipage}{0.5\linewidth}
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}, fill = julia1
        [{$10x$}, fill = julia1
          [{$3x$}, fill = julia1
            [{$1x$}, fill = julia1]
            [{$2x$}, fill = julia2]
          ]
          [{$7x$}, fill = julia3
            [{$3x$}, fill = julia3] 
            [{$4x$}, fill = julia4] 
          ]
        ]
        [{$26x$}, fill = julia5
          [{$11x$}, fill = julia5
            [{$5x$}, fill = julia5] 
            [{$6x$}, fill = julia6] 
          ]
          [{$15x$}, fill = julia7
            [{$7x$}, fill = julia7] 
            [{$8x$}, fill = julia8] 
          ]
        ]
      ]
  \end{forest}
  \end{minipage}%
  \begin{minipage}{0.5\linewidth}
  \begin{forest}
    for tree = {circle, draw, minimum size = 0.8cm}
      [{$36x$}, fill = julia1
        [{$10x$}, fill = julia1
          [{$3x$}, fill = julia1
            [{$1x$}, fill = julia1]
            [{$2x$}, fill = julia1]
          ]
          [{$7x$}, fill = julia1
            [{$3x$}, fill = julia1] 
            [{$4x$}, fill = julia1] 
          ]
        ]
        [{$26x$}, fill = julia2
          [{$11x$}, fill = julia2
            [{$5x$}, fill = julia2] 
            [{$6x$}, fill = julia2] 
          ]
          [{$15x$}, fill = julia2
            [{$7x$}, fill = julia2] 
            [{$8x$}, fill = julia2] 
          ]
        ]
      ]
  \end{forest}
  \end{minipage}
  \caption{Addition of eight floating point numbers across eight (\textbf{left}) 
  and two (\textbf{right}) machines with a guarantee on SPI. Each machine is 
  represented by a different colour. In both cases, the final result can be 
  represented by the same expression $((1x + 2x) + (3x + 4x)) + ((5x + 6x) + (7x + 8x))$.}
  \label{fig:reduction_tree_colour}
\end{figure*}

### 2025-06-26 workshop/introduction.tex
replace PT appendix with nutshell section

Pigeons.jl\footnote{Documentation available at \url{https://pigeons.run/dev/}. 
Public repository: \url{https://github.com/Julia-Tempering/Pigeons.jl}.}
enables users to sample efficiently from challenging probability distributions.
To achieve this, Pigeons.jl provides a multi-threaded and distributed implementation
of non-reversible \emph{parallel tempering}
(PT; \citealp{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity}), 
a state-of-the-art MCMC algorithm (see \cref{sec:PT} for an outline of the method).
Crucially, Pigeons.jl follows the tuning guidelines described in \citet{syed2021nrpt} 
to provide a hyperparameter-free implementation.
Pigeons.jl's simple user interface facilitates running PT
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1,000 processes running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Importantly, Pigeons comes with guarantees on \emph{strong parallelism invariance} (SPI), 
wherein the output for a given seed is \emph{identical} irrespective of the number 
of threads or processes. Such a level of reproducibility is rare in distributed 
software but of great use for the purposes of debugging in the context of sampling 
algorithms, which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

\section{Overview of Pigeons.jl}
A typical instance of the class of problems that Pigeons.jl tackles can be described by
a probability density (or mass) function
%\footnote{This density may also be a probability mass function 
%(discrete variables). We also allow for a combination of discrete/continuous variables.} 
$\pi(x)$ called the \emph{target}. In many problems, e.g. in Bayesian statistics, the 
density $\pi$ is typically known only up to a normalizing constant, 
\[
\label{eq:normalizing_constant}
  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
\]  
where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
Pigeons.jl takes as input the function $\gamma$.

The output of Pigeons.jl can be used for two main tasks:
\begin{enumerate}
    \item Approximating integrals of the form $\int f(x) \pi(x) \, \dee x$.  
    For example, the choice $f(x) = x$ computes the mean and 
    $f(x) = I[x \in A]$ computes the probability of $A$ under $\pi$,
    where $I[\cdot]$ denotes the indicator function.

    \item Approximating the value of the normalization constant $Z$. For 
    example, in Bayesian statistics, this corresponds to the 
    marginal likelihood, which can be used for model selection. 
\end{enumerate}
The Pigeons package particularly shines compared to traditional sampling approaches in the 
following scenarios:
\begin{itemize}
    \item When the target density $\pi$ is challenging due to a complex structure 
    (e.g., high-dimensional, multi-modal, etc.).

    \item When the user needs not only $\int f(x) \pi(x) \, \dee x$ but also
    the normalizing constant $Z$. 
    Many existing tools focus on the former and struggle or fail to do the latter.

    \item When the target distribution $\pi$ is defined over a non-standard space, 
    e.g. a combinatorial object such as a phylogenetic tree.  
\end{itemize}

>>>>

\begin{figure}[t]
    \centering
    \includegraphics[width=0.7\linewidth]{../img/speedup_distributed.pdf}
%    \hspace{4em}
%    \includegraphics[width=0.33\textwidth]{../img/speedup_parallel.pdf}
    \caption{
        Speedup relative to single-thread execution for Pigeons.jl in the distributed 
        (MPI) setting. Error bars denote approximate 95\% confidence intervals.
        The blue dotted line represents a hypothetical speedup equal to the 
        number of processes. The red dashed line indicates a speedup of one for reference.
    }
    \label{fig:PT_scaling}
\end{figure}

Pigeons.jl\footnote{Documentation available at \url{https://pigeons.run/dev/}. 
Public repository: \url{https://github.com/Julia-Tempering/Pigeons.jl}.}
enables users to sample efficiently from challenging probability distributions.
To achieve this, Pigeons.jl provides a multi-threaded and distributed implementation
of non-reversible \emph{parallel tempering}
(PT; \citealp{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity}), 
a state-of-the-art MCMC algorithm.
Crucially, Pigeons.jl follows the tuning guidelines described in \citet{syed2021nrpt} 
to provide a hyperparameter-free implementation.
Pigeons.jl's simple user interface facilitates running PT
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1,000 processes running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Importantly, Pigeons comes with guarantees on \emph{strong parallelism invariance} (SPI), 
wherein the output for a given seed is \emph{identical} irrespective of the number 
of threads or processes. Such a level of reproducibility is rare in distributed 
software but of great use for the purposes of debugging in the context of sampling 
algorithms, which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

### 2025-06-26 workshop/new_developments.tex
replace PT appendix with nutshell section

\paragraph{autoMALA}
Since version v0.2.0, Pigeons.jl's default explorer for most models in continuous
space has been autoMALA \citep{biron2024automala}, a novel MCMC sampler
based on the Metropolis-adjusted Langevin algorithm (MALA). autoMALA continually
adapts its step size at each iteration based on the local geometry of the target
distribution. Such automatic adjustments are highly desirable in the local exploration 
phase of PT (see \cref{sec:PT} for a review of PT), since otherwise we would 
need to find optimal step sizes for each distribution in the annealing path---a 
prohibitively costly process.

>>>>

\paragraph{autoMALA}
Since version v0.2.0, Pigeons.jl's default explorer for most models in continuous
space has been autoMALA \citep{biron2024automala}, a novel MCMC sampler
based on the Metropolis-adjusted Langevin algorithm (MALA). autoMALA continually
adapts its step size at each iteration based on the local geometry of the target
distribution. Such automatic adjustments are highly desirable in the exploration 
phase, since otherwise we would 
need to find optimal step sizes for each distribution in the annealing path---a 
prohibitively costly process.

### 2025-06-26 workshop/overview.tex
replace PT appendix with nutshell section

\section{Introduction}
In many scientific application domains, the ability to obtain samples from a 
probability distribution is of central interest. 
For instance, sampling methods have been used to discover magnetic polarization 
in the black hole of galaxy M87 \cite{akiyama2021seven}
and to image the Sagittarius A* black hole \cite{akiyama2022first}.
They have also been used to 
model the evolution of single-cell cancer genomes \cite{salehi2021clonal}, 
infer plasma dynamics inside nuclear fusion reactors \cite{gota2021overview}, 
and to identify gerrymandering in Georgia's 2021 congressional districting plan 
\cite{zhao2022mathematically}.
Similarly, evaluating high-dimensional integrals or sums over complicated 
combinatorial spaces are related tasks that can also be solved with sampling 
via Markov chain Monte Carlo (MCMC) methods. 
However, such calculations can often be bottlenecks in the scientific process, with 
simulations that can last days or even weeks to finish.

\medskip
Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0, used in the example scripts below, is compatible with Julia 1.8+.}
enables users to sample efficiently from challenging probability distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple user interface facilitates performing such computations 
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1000 machines running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Further, Pigeons comes with guarantees on strong parallelism invariance wherein 
the output for a given seed is \emph{identical} irrespective 
of the number of threads or processes. 
Such a level of reproducibility is rare in distributed software but of 
great use for the purposes of debugging in the context of sampling algorithms, 
which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

\subsection{Latest documentation}

More in-depth documentation and additional examples can be found at: {\renewcommand{\UrlFont}{\ttfamily\normalsize}\url{https://pigeons.run/}}.

\subsection{Problem formulation}
We describe the class of problems that can be approached using Pigeons.jl.
Let $\pi(x)$ denote a probability density\footnote{This density may 
also be a probability mass function 
(discrete variables). We also allow for a combination of discrete/continuous variables.} 
called the \emph{target}. 
In many problems, e.g. in Bayesian statistics, the density $\pi$ is typically 
known only up to a normalizing constant, 
\[
\label{eq:normalizing_constant}
  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
\]  
where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
Pigeons.jl takes as input the function $\gamma$.

\medskip 
The output of Pigeons.jl can be used for two main tasks:
\begin{enumerate}
    \item Approximating integrals of the form $\int f(x) \pi(x) \, \dee x$.  
    For example, the choice $f(x) = x$ computes the mean and 
    $f(x) = I[x \in A]$ computes the probability of $A$ under $\pi$,
    where $I[\cdot]$ denotes the indicator function.

@@

\subsection{What is of interest to the general Julia developer?}
Ensuring code correctness at the intersection of randomized, parallel, and 
distributed algorithms is a challenge. To address this challenge, we designed Pigeons.jl 
based on a principle that we refer to as \textit{strong parallelism invariance}
(SPI).
Namely, the output of Pigeons.jl is  invariant to the number of machines 
and/or threads.
Without explicitly keeping SPI in mind during software construction, 
the (random) output of the algorithm is only guaranteed to have the same distribution.
This is a much weaker guarantee that, in particular, makes debugging difficult.
However, with our notion of SPI we make debugging and 
software validation considerably easier. This is because the developer can 
first focus on the fully serial randomized algorithm, and then use it as an 
easy-to-compare gold-standard reference for parallel/distributed implementations. 
This strategy is used extensively in Pigeons.jl to ensure correctness. In contrast, 
testing equality in distribution, while possible (e.g., see \cite{geweke2004getting}), 
incurs additional false negatives due to statistical error.

\medskip 
The general Julia developer will be interested in: 
\begin{itemize}
    \item The main causes of a violation of strong parallelism invariance that we have identified 
    (\cref{sec:PI_causes})---some of which are specific to Julia---and how we address 
    them in Pigeons.jl.

    \item The SplittableRandoms.jl\footnote{\url{https://github.com/Julia-Tempering/SplittableRandoms.jl}}
    package that was developed by our team 
    to achieve strong parallelism invariance in Pigeons.jl (\cref{sec:splittable_randoms}).
\end{itemize}

>>>>

\section{Overview of Pigeons.jl}\label{sec:overview}

Pigeons.jl exploits the fact that the exploration phase in PT is embarrassingly parallel
to provide a multithreaded and distributed Julia implementation of the algorithm.

In many problems, e.g. in Bayesian 
statistics, the target $\pi$ is typically known only up to a normalizing constant, 
\[
\label{eq:normalizing_constant}
  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
\]  
where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
Pigeons.jl takes as input the function $\gamma$.

The output of Pigeons.jl can be used for two main tasks:
\begin{enumerate}
    \item Approximating integrals of the form $\int f(x) \pi(x) \, \dee x$.  
    For example, the choice $f(x) = x$ computes the mean and 
    $f(x) = I[x \in A]$ computes the probability of $A$ under $\pi$,
    where $I[\cdot]$ denotes the indicator function.

@@
